---
title: "ps1"
output: github_document
date: "2023-09-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading in Required Packages

```{r}
library(tidyverse)
```

## Loading in and Cleaning Data

```{r}
five_yr_pop_data <- read_delim("/Users/kenjinchang/github/phd-problemsets/data/idb5yr.txt")
pyramid_data <- five_yr_pop_data %>% rename("YR"="#YR") %>% filter(YR==2022) %>% filter(GEO_ID=="W140000WOCM"|GEO_ID=="W140000WOKE"|GEO_ID=="W140000WONG"|GEO_ID=="W140000WOZA") %>% select(YR,GEO_ID,FPOP0_4,FPOP5_9,FPOP10_14,FPOP100_,FPOP15_19,FPOP20_24,FPOP25_29,FPOP30_34,FPOP35_39,FPOP40_44,FPOP45_49,FPOP50_54,FPOP55_59,FPOP60_64,FPOP65_69,FPOP70_74,FPOP75_79,FPOP80_84,FPOP85_89,FPOP90_94,FPOP95_99,MPOP0_4,MPOP5_9,MPOP10_14,MPOP100_,MPOP15_19,MPOP20_24,MPOP25_29,MPOP30_34,MPOP35_39,MPOP40_44,MPOP45_49,MPOP50_54,MPOP55_59,MPOP60_64,MPOP65_69,MPOP70_74,MPOP75_79,MPOP80_84,MPOP85_89,MPOP90_94,MPOP95_99) %>% pivot_longer(cols=c(FPOP0_4,FPOP5_9,FPOP10_14,FPOP15_19,FPOP20_24,FPOP25_29,FPOP30_34,FPOP35_39,FPOP40_44,FPOP45_49,FPOP50_54,FPOP55_59,FPOP60_64,FPOP65_69,FPOP70_74,FPOP75_79,FPOP80_84,FPOP85_89,FPOP90_94,FPOP95_99,FPOP100_,MPOP0_4,MPOP5_9,MPOP10_14,MPOP100_,MPOP15_19,MPOP20_24,MPOP25_29,MPOP30_34,MPOP35_39,MPOP40_44,MPOP45_49,MPOP50_54,MPOP55_59,MPOP60_64,MPOP65_69,MPOP70_74,MPOP75_79,MPOP80_84,MPOP85_89,MPOP90_94,MPOP95_99),
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                  names_to="age.range",
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                  values_to="population") %>%
  mutate(sex=case_when(str_detect(age.range,"^F")~"F",TRUE~"M")) %>%
  mutate(age.range=str_remove_all(age.range,"MPOP|FPOP"))
pyramid_data %>% head(6)
```

ISO codes for Cameroon, Kenya, Nigeria, and South Africa are CM, KE, NG, and ZA, respectively.The values in the 'GEO_ID' column therefore correspond to the following string of characters: 'W140000WOCM', 'W140000WOKE', 'W140000WONG', and 'W140000WOZA'.

### Pyramid Construction

The IDB dataset is now *mostly* in a format we can use to generate population-by-age pyramids. The only thing left to do is to select rows of observations according to country. 

```{r}
cameroon_pyramid_data <- filter(pyramid_data,GEO_ID=="W140000WOCM")
cameroon_pyramid_data %>% head(6)
```

```{r}
kenya_pyramid_data <- filter(pyramid_data,GEO_ID=="W140000WOKE")
kenya_pyramid_data %>% head(6)
```
```{r}
nigeria_pyramid_data <- filter(pyramid_data,GEO_ID=="W140000WONG")
nigeria_pyramid_data %>% head(6)
```
```{r}
southafrica_pyramid_data <- filter(pyramid_data,GEO_ID=="W140000WOZA")
southafrica_pyramid_data %>% head(6)
```

Now, with these subsets of the 'pyramid_data' stored in our environment, we can finally generate the corresponding pyramid plots.

## Population by Age Pyramids

```{r}
cameroon_pyramid_data %>% mutate(population=ifelse(sex=="M",population*(-1),
                                                      population*(1))) %>% 
  ggplot(aes(x=age.range,y=population,fill=sex)) + 
  geom_bar(stat="identity") + 
  scale_x_discrete(limits=c("0_4","5_9","10_14","15_19","20_24","25_29","30_34","35_39","40_44","45_49","50_54","55_59","60_64","65_69","70_74","75_79","80_84","85_89","90_94","95_99","100_")) +
  coord_flip() +
  ggtitle("Cameroon")
```

```{r}
kenya_pyramid_data %>% mutate(population=ifelse(sex=="M",population*(-1),
                                                      population*(1))) %>% 
  ggplot(aes(x=age.range,y=population,fill=sex)) + 
  geom_bar(stat="identity") + 
  scale_x_discrete(limits=c("0_4","5_9","10_14","15_19","20_24","25_29","30_34","35_39","40_44","45_49","50_54","55_59","60_64","65_69","70_74","75_79","80_84","85_89","90_94","95_99","100_")) +
  coord_flip() +
  ggtitle("Kenya")
```

```{r}
nigeria_pyramid_data %>% mutate(population=ifelse(sex=="M",population*(-1),
                                                      population*(1))) %>% 
  ggplot(aes(x=age.range,y=population,fill=sex)) + 
  geom_bar(stat="identity") + 
  scale_x_discrete(limits=c("0_4","5_9","10_14","15_19","20_24","25_29","30_34","35_39","40_44","45_49","50_54","55_59","60_64","65_69","70_74","75_79","80_84","85_89","90_94","95_99","100_")) +
  coord_flip() +
  ggtitle("Nigeria")
```


```{r}
southafrica_pyramid_data %>% mutate(population=ifelse(sex=="M",population*(-1),
                                                      population*(1))) %>% 
  ggplot(aes(x=age.range,y=population,fill=sex)) + 
  geom_bar(stat="identity") + 
  scale_x_discrete(limits=c("0_4","5_9","10_14","15_19","20_24","25_29","30_34","35_39","40_44","45_49","50_54","55_59","60_64","65_69","70_74","75_79","80_84","85_89","90_94","95_99","100_")) +
  coord_flip() +
  ggtitle("South Africa")
```

We can see that the age distributions across the four investigated countries vary considerably. South Africa, in particular, stands out, as there are more people aged 25-29, 30-34, 35-39, and 40-44 than people aged 20-24.A similar emergent pattern is true for the Kenyan population, with there being more people aged 10-14 than those aged 5-9. However, there are some notable consistencies in the distributions across gender, with there being more females at older age bins relative to males.

## Dependency Ratios

For this section, we will need to reconfigure the data so that it can serve our calculations. 

```{r}
five_yr_pop_data <- read_delim("/Users/kenjinchang/github/phd-problemsets/data/idb5yr.txt")
dependency_data <- five_yr_pop_data %>% rename("YR"="#YR") %>% filter(YR==1990|YR==2000|YR==2010|YR==2020|YR==2030|YR==2040|YR==2050) %>% filter(GEO_ID=="W140000WOCM"|GEO_ID=="W140000WOKE"|GEO_ID=="W140000WONG"|GEO_ID=="W140000WOZA")
dependency_data <- dependency_data %>% select(YR,GEO_ID,POP0_4,POP5_9,POP10_14,POP15_19,POP20_24,POP25_29,POP30_34,POP35_39,POP40_44,POP45_49,POP50_54,POP55_59,POP60_64) %>% rowwise() %>% mutate(dependents=sum(POP0_4,POP5_9,POP10_14)) %>% mutate(working=sum(POP15_19,POP20_24,POP25_29,POP30_34,POP35_39,POP40_44,POP45_49,POP50_54,POP55_59,POP60_64)) %>% select(YR,GEO_ID,dependents,working) %>% mutate(dependency.ratio=dependents/working)
dependency_data
```

With this all now configured, we can graph the dependency ratios.

```{r}
ggplot(dependency_data,aes(x=YR,y=dependency.ratio,color=GEO_ID)) + 
  geom_point() +
  geom_smooth(se=FALSE) 
```

Based on the above, we can see that, overall, youth dependency is decreasing, meaning the population aged 15-64, i.e., the denominator, is increasing at a faster rate than the the population aged 0-15, i.e., the numerator. We can see that this is expected to happen more suddenly in some placed than others, e.g., the downward slope demonstrated for Kenya relative to the softer slope for Cameroon.

### Fertility Rates

Hey, Matt. Had a little trouble using the STATcompiler tool - may be a browser thing. Will update this repo/doc once I figure out the problem. 
