---
title: "ps3"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## comm 6750, lab 3: getting network data into R and doing simple calculations

### required packages

```{r}
library(sna)
```

### reading in and converting senator data

```{r}
senator_data <- read.csv("/Users/kenjinchang/github/phd-problemsets/data/nys_senators_edgelist.csv")
senator_data_asmat <- as.matrix(senator_data)
```

### creating network



```{r}
senator_data_asnet <- as.network(senator_data_asmat)
```

### network visualization

```{r}
gplot(senator_data_asnet)
```

### calculating density

```{r}
gden(senator_data_asnet)
```

### calculating reciprocity

```{r}
grecip(senator_data_asnet,measure="edgewise")
```

this represents the portion of edges that are reciprocated.

```{r}
grecip(senator_data_asnet)
```

while this represents the portion of dyads that are reciprocated.

### repeating steps using igraph

```{r}
detach("package:sna",unload=TRUE)
```

```{r}
library(igraph)
```

```{r}
dim(senator_data_asmat)
```

Issues with non-square matrix - will resolve later after researching function parameters.


graph_from_adjacency_matrix(adjmatrix=senator_data_asmat, mode="directed",diag=FALSE)


